Chess EDA

### 2023-03-21 owen/01_eda.qmd
tidying up the models and the code

---
title: "Chess EDA"
format: html
bibliography: refs.bib
---

@@

We first wish to investigate the properties of the scraped data.

@@

We see that the vast majority of games in this dataset 
are Standard, with the regular rules, etc. Similarly, the most
commonly played time controls is 180+0,
which corresponds to 3 minute blitz games with no time increment.

Of standard games, we see that the majority are rated blitz and bullet games.
These correspond to randomly playing available players on the site 
at a time. This differs from tournaments, which make up a sizeable proportion
of standard games also.

```{r}
table(lichess_data$Variant)

head(sort(table(lichess_data$TimeControl), decreasing = TRUE))

## this looks to be seconds + incrememnt, makes the most sense 
## could just take common blitz formats, 3 mins and 5 mins,
## would be 50k games still

head(sort(table(lichess_data$Event), decreasing = TRUE))

lichess_data %>% filter(Variant == "Standard") %>% 
  group_by(Event) %>% 
  count() %>% 
  arrange(-n)

@@

Get all games from a specific player in a specific format,
to see that `RatingDiff` tracks how
the rating changes. For the White/Black player based
on the result of the current game.

```{r}
lichess_data %>% filter(TimeControl == "180+0") %>% 
  filter(Event == "Rated Blitz game") %>% 
  arrange(UTCDate, UTCTime) %>% 
  mutate(peng_elo = ifelse(White == "penguingim1", WhiteElo, BlackElo)) %>% 
  mutate(peng_change = ifelse(White == "penguingim1", WhiteRatingDiff,
                              BlackRatingDiff)) %>% 
  select(UTCDate, UTCTime, peng_elo, peng_change) %>%
  mutate(peng_elo = as.numeric(peng_elo)) %>% 
  mutate(peng_change = as.numeric(peng_change)) %>% 
  mutate(ELO_change = peng_elo + peng_change) %>% 
  head(n = 10)

### so RatingDiff gives you the starting ELO for the next game

@@

- Casual games don't change the rating, so do we want to remove them from 
the data entirely? Would likely have a different role in winner effects,
etc.

For all current analysis we will restrict to 1 minute rated bullet games,
the most commonly played format in this dataset, with 13k games compared
to 12k games of 30 second bullet and 12k games of 3 minute blitz.

```{r}
## get simpler data

simple_data <- lichess_data %>% 
  filter(Event %in% c("Rated Bullet game", "Rated Blitz game"))

@@

bullet_60 <- lichess_data %>% 
  filter(Event == "Rated Bullet game") %>% 
  filter(TimeControl == "60+0")

# bullet_60

# head(sort(table(bullet_60$White), decreasing = TRUE), 20)

## have a few different players with good number of games
## get list of all players, fit some sort of model using all players
## who play enough games
```

@@

```{r}
head(sort(table(c(bullet_60$White, bullet_60$Black)), decreasing = TRUE), 25)

## top 25 players possibly enough then, consider all games they are involved
## in

table(bullet_60$Result)

@@

## Multilevel Logistic Regression

Similar to @ding2021nhl, we wish to fit a multilevel logistic 
regression to see if we can identify individual 
effects for chess players in this data.

To do this, we will consider the current history of an individual player
as their win ratio in their past $n$ games, where we will investigate
the sensitivity of these results to different choices of $n$.

We will identify focal players as those who were used to query the API,
i.e the `Username` for each game. We only consider Rated 60 second bullet games
here.

```{r, past-n-results}
#| message: false

@@

get_hist <- function(user, games, prev_n) {
  hist_games <- games %>% 
    filter(White == user | Black == user) %>% 
    arrange(UTCDate, UTCTime) %>% 
    mutate(focal_white = ifelse(Username == White, 1, 0)) %>% 
    select(White:BlackElo, focal_white) %>% 
    mutate(focal_result = case_when(
      (focal_white == 1 & Result == "1-0") ~ 1,
      (focal_white == 0 & Result == "0-1") ~ 1,
      (Result == "1/2-1/2") ~ 0.5,
      .default = 0
      )) %>% 
    mutate(focal_win_prop = c(rep(0, prev_n - 1), 
                              roll_mean(focal_result, n = prev_n)))

@@

```{r}
## want to use focal players not all players to get these lists to be the same
a <- head(sort(table(c(bullet_60$Username)), decreasing = TRUE), 17)
## if don't use all focal players here then removing data because they won't be
## involved

top_players <- names(a)

@@

tidy_games %>% 
  mutate(focal_player = ifelse(focal_white == 1, White, Black)) %>% 
  group_by(focal_player) %>% 
  count() %>% 
  arrange(-n)

## why does this have 17 players rather than 25 then?
## why are some missing?

@@

Now need to figure out how to feed this into a Stan model.

```{r}
head(tidy_games)

init_data <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, top_players)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% mutate(ave_prop = focal_win_prop - mean(focal_win_prop))

init_data

### will need
### number of focal players
### which focal player is in each row
### their outcome 0 or 1
### their colour white (1) or black (0)
### ranking diff

@@

```{r}

@@

```{r}
stan_file <- "model2.stan"

fit2 <- mod$sample(data = stan_data,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

@@

```{r}
stan_file <- "model3.stan"

fit3 <- mod$sample(data = stan_data,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit3$summary()
mcmc_hist(fit3$draws(c("beta", "mu", "tau")))

@@

```{r}

stan_data_ave <- list(N = nrow(init_data),
                  J = length(top_players),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$ave_prop)

@@

```{r}
stan_file <- "model2.stan"

@@

```{r}
stan_file <- "model3.stan"

@@

```{r}
#| include: false

>>>>

---
title: "Chess EDA"
author: Owen G. Ward
format: 
  html:
    toc: true
    toc-location: left
bibliography: refs.bib
---

@@

We first wish to investigate the properties of the scraped data. We 
focus for now on the Lichess data. We see that the vast majority of games in this dataset 
are Standard Chess, with the regular rules, etc.

@@

```{r}
#| echo: true
## Variants of Chess in this data
table(lichess_data$Variant)
```
Similarly, the most
commonly played time control, across all formats of chess, is 180+0,
which corresponds to 3 minute blitz games with no time increment.

```{r}
#| echo: true
head(sort(table(lichess_data$TimeControl), decreasing = TRUE))
```

Of standard games, we see that the majority are rated blitz and bullet games.
These correspond to randomly playing available players on the site 
at a given time. This differs from tournaments, which make up a sizeable proportion
of standard games also (and which might have different dynamics).

```{r}
#| echo: true
head(sort(table(lichess_data$Event), decreasing = TRUE))

@@

```{r}
#| echo: true

lichess_data %>% filter(Variant == "Standard") %>% 
  group_by(Event) %>% 
  count() %>% 
  arrange(-n) %>% 
  head()

@@

The variable `ratingDiff` records how the ELO of both the black and
white player changes based on the outcome of that game.
Here these ELO scores are across the different event types, which
may contain multiple time formats. For example, an ELO for Bullet play will 
change when both 1 minute and 30 second games are played, and also in 
both tournament and rated games.

<!-- Get all games from a specific player in a specific format, -->
<!-- to see that `RatingDiff` tracks how -->
<!-- the rating changes. For the White/Black player based -->
<!-- on the result of the current game. -->

```{r}
#| echo: false
#| eval: false
elo_check <- lichess_data %>% filter(TimeControl == "180+0") %>% 
  filter(Event == "Rated Blitz game") %>% 
  arrange(UTCDate, UTCTime) %>% 
  mutate(peng_elo = ifelse(White == "penguingim1", WhiteElo, BlackElo)) %>% 
  mutate(peng_change = ifelse(White == "penguingim1", WhiteRatingDiff,
                              BlackRatingDiff)) %>% 
  select(UTCDate, UTCTime, peng_elo, peng_change) %>%
  mutate(peng_elo = as.numeric(peng_elo)) %>% 
  mutate(peng_change = as.numeric(peng_change)) %>% 
  mutate(ELO_change = peng_elo + peng_change)

elo_check %>% 
  mutate(diff = peng_elo - lag(ELO_change))

### so RatingDiff gives you the starting ELO for the next game of the same type

elo_check_2 <- lichess_data %>% 
  filter(Event == "Rated Blitz game") %>% 
  arrange(UTCDate, UTCTime) %>% 
  mutate(peng_elo = ifelse(White == "penguingim1", WhiteElo, BlackElo)) %>% 
  mutate(peng_change = ifelse(White == "penguingim1", WhiteRatingDiff,
                              BlackRatingDiff)) %>% 
  select(UTCDate, UTCTime, peng_elo, peng_change) %>% 
  mutate(peng_elo = as.numeric(peng_elo)) %>% 
  mutate(peng_change = as.numeric(peng_change)) %>% 
  mutate(ELO_change = peng_elo + peng_change) %>% 
  mutate(diff = peng_elo - lag(ELO_change))

elo_check_2

## so these don't always agree, because could play arena games, etc

Casual games don't change the rating, so we should probably treat them
differently.
Would likely have a different role in winner effects,
etc. It is possible that Arena and tournament games also have different 
dynamics.

## Initial Models

For the current initial analysis we
will restrict to 1 minute rated bullet games,
the most commonly played format in this dataset, with 13k games compared
to 12k games of 30 second bullet and 12k games of 3 minute blitz.

```{r}
#| echo: true
## get initial dataset to use for analysis

@@

bullet_60 <- lichess_data %>% 
  filter(Event == "Rated Bullet game") %>% 
  filter(TimeControl == "60+0")
```

@@

```{r}
#| echo: true
head(sort(table(c(bullet_60$White, bullet_60$Black)), decreasing = TRUE), 25)

## results of these games
table(bullet_60$Result)

@@

### Multilevel Logistic Regression

Similar to @ding2021nhl, we wish to fit a multilevel logistic 
regression to see if we can identify individual 
effects for chess players in this data.
We will incorporate fixed effects for the player effects of the focal
player in these games, along

We will identify focal players as those who were used to query the API,
i.e the `Username` for each game. We only consider Rated 60 second bullet games
here, and use the top 18 players, who all play at least 100 games
of 1 minute rated bullet chess in this dataset.

We will consider the current history of an individual player
as their win ratio in their past $n$ games, where we will investigate
the sensitivity of these results to different choices of $n$.

As a small percentage (less than 4%) of these games result in 
a draw, we will use these draws to get each players current win history,
before then removing these from the logistic regression model.
Initially we will choose the previous $$n=10$$ games for each player to
compute the win ratio, or the average of all current games if
the historic number of games played is less than 10.

```{r get historic win ratio}

get_hist <- function(user, games, prev_n) {
  hist_games <- games %>% 
    filter(White == user | Black == user) %>% 
    arrange(UTCDate, UTCTime) %>% 
    mutate(focal_white = ifelse(Username == White, 1, 0)) %>% 
    select(White:BlackElo, focal_white) %>% 
    mutate(focal_result = case_when(
      (focal_white == 1 & Result == "1-0") ~ 1,
      (focal_white == 0 & Result == "0-1") ~ 1,
      (Result == "1/2-1/2") ~ 0.5,
      .default = 0
      )) %>% 
    mutate(focal_win_prop = c(cumsum(focal_result[1:(prev_n - 1)])/(1:(prev_n -1)), 
                              roll_mean(focal_result, n = prev_n)))
  hist_games
}

```{r, past-n-results}
#| message: false
#| eval: false

@@

### modified this so that it calculates rolling averarge of previous n
### games, or all games, when number of previous games is less than n

get_hist <- function(user, games, prev_n) {
  hist_games <- games %>% 
    filter(White == user | Black == user) %>% 
    arrange(UTCDate, UTCTime) %>% 
    mutate(focal_white = ifelse(Username == White, 1, 0)) %>% 
    select(White:BlackElo, focal_white) %>% 
    mutate(focal_result = case_when(
      (focal_white == 1 & Result == "1-0") ~ 1,
      (focal_white == 0 & Result == "0-1") ~ 1,
      (Result == "1/2-1/2") ~ 0.5,
      .default = 0
      )) %>% 
    mutate(focal_win_prop = c(cumsum(focal_result[1:(prev_n - 1)])/(1:(prev_n -1)), 
                              roll_mean(focal_result, n = prev_n)))

@@

We can see the distribution of the win ratio for these 18 players across all
games in this sample.

```{r}
#| message: false

## want to use focal players not all players to get these lists to be the same
focal_players <- head(sort(table(c(bullet_60$Username)), decreasing = TRUE), 18)
## if don't use all focal players here then removing data because they won't be
## involved

top_players <- names(focal_players)

@@

tidy_games %>% 
  ggplot(aes(focal_win_prop)) +
  geom_histogram() +
  labs(x = "Win Ratio", title = "Previous 10 Games")

@@

Removing the drawn games, we have 12500 games involving these 18 focal players.

```{r organize data for stan}

init_data <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, top_players)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% 
  mutate(ave_prop = focal_win_prop - mean(focal_win_prop)) %>% 
  filter(focal_result != 0.5)

nrow(init_data)

@@

#### STAN Model

Now we want to fit this multilevel logistic regression. Letting
$y_{ij}$ be the result of the $i$-th game by the $j$-th focal player,
then the model we will consider is.

---

```{r}
#| eval: false

@@

```{r}
#| eval: false
stan_file <- "model2.stan"

mod2 <- cmdstan_model(stan_file)

fit2 <- mod2$sample(data = stan_data,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

@@

```{r}
#| eval: false
stan_file <- "model3.stan"

mod3 <- cmdstan_model(stan_file)

fit3 <- mod3$sample(data = stan_data,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit3$summary()
mcmc_hist(fit3$draws(c("beta", "mu", "tau", "gamma1", "gamma2")))

@@

```{r}
#| eval: false
stan_data_ave <- list(N = nrow(init_data),
                  J = length(top_players),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$ave_prop)

@@

```{r}
#| eval: false
stan_file <- "model2.stan"

@@

```{r}
#| eval: false
stan_file <- "model3.stan"

@@

```{r}
#| include: false
#| eval: false
